### 2014-02-21 thesis-proposal/parsing.tex
Essentially done Sec. 4, beginning of Sec. 5, rough draft of end of Sec. 5

\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
    Empty of Unit
  | Text of String
  | Seq of HTML * HTML 
  | BodyElement of Attributes * HTML
  | StyleElement  of Attributes * CSS
  | ...
  metadata = new : HasTSL
    val parser : Parser = ~
SGRM      start -> ':body'= start>
        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
SGRM      start -> ':style'= EXP>
        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
SGRM      start -> '{'= EXP '}'=EGRM
        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.53\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ts : TokenStream) : (Exp * 
    TokenStream)
  metadata = new : HasTSL
    val parser : Parser = (* parser generator *)      
casetype Type
    Var of ID
  | Arrow of Type * Type
  | Prod of Type * Type
  metadata = new : HasTSL
    val parser : Parser = (* type quasiquotes *)
\end{lstlisting}
\end{subfigure}
\begin{subfigure}[t]{.55\textwidth}
\begin{lstlisting}[linewidth=.5\textwidth]
casetype Exp 
    Var of ID
  | Lam of ID * Type * Exp
  | Ap of Exp * Exp
  | ProdIntro of Exp * Exp
  | CaseIntro of Type * ID * Exp
  ... 
  | FromTS of Tokenstream
  | Literal of TokenStream
  | Error of ErrorMessage
  metadata = new : HasTSL
    val parser : Parser = (* quasiquotes *)
\end{lstlisting}
\end{subfigure}
\caption{Some of the types included in the Wyvern prelude.}
\vspace{-10px}
\label{f-builtins}
%\label{fig:typeParser}
\end{figure}

@@

\subsection{Remaining Tasks and Timeline}
The paper we have submitted to ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar. 7th). If accepted, final revisions on the paper text are due on May 12th and so we will largely do these tasks after the OOPSLA deadline in late March. If not accepted, we will aim to resubmit to OOPSLA.
\begin{enumerate}
%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints. This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example). We will add these simple rules to the grammar.
\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory. This might require slightly restricting the recursion in the rule for literals, so that induction is well-founded.
\item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this. In the paper, the formalism is meant to be expository, so we will likely leave this additional complexity for the tech report.
\item We must further consider aspects of hygiene:
\begin{enumerate}
\item We do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}. It may be possible to prevent shadowing by making it impossible for variables to be introduced into interpolated Wyvern expressions directly (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
\item In macro systems like Scheme's, free variables in generated ASTs refer to their values within the macro definition, rather than where they are inserted. To support this, we have introduced the \li{valAST} operator. Using it is a bit verbose and awkward -- every free variable in a quasiquote must be inside \li{valAST}. We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using a very simple extension to the interpolation mechanism that allows interpolation of single variables, rather than whole expressions, and plan to write up how this works.
\end{enumerate}
%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
%\item The current implementation of Wyvern does not include all of these mechanisms as described. Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal? I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
\end{enumerate}
%
%\subsection{TODO}
%- Figure out separate compilation (e.g. of CSS)

>>>>

\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
    Empty of Unit
  | Text of String
  | Seq of HTML * HTML 
  | BodyElement of Attributes * HTML
  | StyleElement  of Attributes * CSS
  | ...
  metadata = new : HasTSL
    val parser : Parser = ~
SGRM      start -> ':body'= start>
        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
SGRM      start -> ':style'= EXP>
        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
SGRM      start -> '{'= EXP '}'=EGRM
        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.53\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ts : TokenStream) : (Exp * 
    TokenStream)
  metadata = new : HasTSL
    val parser : Parser = (* parser generator *)      
casetype Type
    Var of ID
  | Arrow of Type * Type
  | Prod of Type * Type
  metadata = new : HasTSL
    val parser : Parser = (* type quasiquotes *)
\end{lstlisting}
\end{subfigure}
\begin{subfigure}[t]{.55\textwidth}
\begin{lstlisting}[linewidth=.5\textwidth]
casetype Exp 
    Var of ID
  | Lam of ID * Type * Exp
  | Ap of Exp * Exp
  | ProdIntro of Exp * Exp
  | CaseIntro of Type * ID * Exp
  ... 
  | FromTS of Tokenstream
  | Error of ErrorMessage
  metadata = new : HasTSL
    val parser : Parser = (* quasiquotes *)
\end{lstlisting}
\end{subfigure}
\caption{Some of the types included in the Wyvern prelude.}
\vspace{-10px}
\label{f-builtins}
%\label{fig:typeParser}
\end{figure}

@@

\subsection{Remaining Tasks and Timeline}
The paper we have submitted to ECOOP 2014 has received quite positive reviews, so we anticipate that it will be accepted (notifications are sent on Mar. 7th). If accepted, final revisions on the paper text are due on May 12th and so we will largely do these tasks after the OOPSLA deadline in late March. If not accepted, we will aim to resubmit to OOPSLA.
\begin{enumerate}
%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints. This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example). We will add these simple rules to the grammar.
\item We must write down the full formal semantics (including rules that we omitted for concision in the paper draft) in a technical report, as well as provide more complete proofs of the metatheory. This might require slightly restricting the recursion in the rule for literals, so that induction is well-founded.
\item Our current static semantics does not support mutually recursive type declarations, but this is necessary for our prelude to typecheck, so we will add support for this. In the paper, the formalism is meant to be expository, so we will likely leave this additional complexity for the tech report.
\item We must further consider aspects of hygiene:
\begin{enumerate}
\item We do not yet have a clean mechanism for preventing unintentional variable \emph{shadowing}, only unintentional variable \emph{capture}. It may be possible to prevent shadowing by preventing variables from leaking into interpolated Wyvern expressions (so that function application is the only way to pass data from a TSL to Wyvern code, as in the \li{Parser} TSL above).
\item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted. To support this, we have introduced the \li{valAST} operator. Using it is a bit verbose and awkward -- every free variable \li{x} in a quasiquote must instead be replaced with an interpolated term of the form \li{valAST(x)} for the resulting \li{Exp} to meaningful as the return value of the \li{parse} function. We believe we can insert \li{valAST} automatically from within the definition of the TSL for \li{Exp} using single variable (rather than whole expression) interpolation.
\end{enumerate}
%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
%\item The current implementation of Wyvern does not include all of these mechanisms as described. Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal? I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
\end{enumerate}
%
%\subsection{TODO}
%- Figure out separate compilation (e.g. of CSS)
